A 2-Norm Condition Number for B\'{e}zier Curve Intersection

### 2019-10-18 doc/paper.tex
Updates based on 2nd round of review (part 1).
- Replacing " then" with ", " as a pause
- Using bold 0 (i.e. vector) when specializing dG/d(delta) to delta = 0
- Replacing "It's" with "It is"
- Adding "a sum of" to sentence about terms in Theorem 4.1
- Removing "of" from "The factor of epsilon"
- Replacing "we'd" with "we would"
- Using "Coincidence" instead of "Coincident" in section 5.4
  heading
- Adding "we" in "Again we have a condition number"

These changes are only in `paper.tex`, I have avoided checking in
`paper.pdf` until all changes are complete (since a binary file creates
a somew

Consider a smooth function \(F: \reals^n \longrightarrow \reals^n\)
with Jacobian \(F_{\bm{x}} = J\). We want to consider a special class of
functions of the form \(F\left(\bm{x}\right) = \sum_j c_j
\phi_j\left(\bm{x}\right)\) where the basis
functions \(\phi_j\) are also smooth functions
\(\reals^n \longrightarrow \reals^n\)
and each \(c_j \in \reals\). We want to consider the effects on a root
\(\bm{\alpha} \in \reals^n\) of a perturbation in one of the
coefficients \(c_j\). We examine the perturbed function
\begin{equation}
G\left(\bm{x}, \delta\right) = F\left(\bm{x}\right) +
\delta \phi_j\left(\bm{x}\right).
\end{equation}
Since \(G\left(\bm{\alpha}, 0\right) = \bm{0}\), if \(J^{-1}\) exists at
\(\bm{x} = \bm{\alpha}\) then
the implicit function theorem tells us that we can define
\(\bm{x}\) via
\begin{equation}
G\left(\bm{x}\left(\delta\right), \delta\right) = \bm{0}.
\end{equation}
Taking the derivative with respect to \(\delta\) we find that
\(\bm{0} = G_{\bm{x}} \bm{x}' + G_{\delta}\). Plugging in
\(\delta = 0\) we find that \(0 = J\left(\bm{\alpha}\right) \bm{x}' +
\phi_j\left(\bm{\alpha}\right)\), hence we
conclude that
\begin{equation}
\bm{x}\left(\delta\right) = \bm{\alpha} - J^{-1}\left(\bm{\alpha}\right)
  \phi_j\left(\bm{\alpha}\right) \delta + \bigO{\delta^2}.
\end{equation}
This gives
\begin{equation}
\frac{\left \lVert J^{-1}\left(\bm{\alpha}\right)
  \phi_j\left(\bm{\alpha}\right) \right \rVert}{
  \left \lVert \bm{\alpha} \right \rVert}.
\end{equation}
as the relative condition number for a perturbation in \(c_j\).
By considering perturbations in \emph{all} of the coefficients:
\(\left|\delta_j\right| \leq \eps \left|c_j\right|\), a similar analysis
gives a root function
\begin{equation}
\bm{x}\left(\delta_0, \ldots, \delta_n\right) = \bm{\alpha} -
  J^{-1}\left(\bm{\alpha}\right) \sum_{j = 0}^n \delta_j
  \phi_j\left(\bm{\alpha}\right) + \bigO{\eps^2}.
\end{equation}
With this, we can define a root condition number

@@

In \cite[Chapter~25, Section 25.4]{Higham2002} a similar definition is
given. Instead of bounding the perturbations component-wise, it bounds
the entire perturbation vector \(\bm{\delta}\)
\begin{equation}
  \lim_{\eps \to 0} \left(\sup_{\left \lVert \bm{\delta} \right \rVert \leq
  \eps \left \lVert \bm{c} \right \rVert} \frac{\left \lVert
  J^{-1}\left(\bm{\alpha}\right) \sum_j \delta_j
  \phi_j\left(\bm{\alpha}\right) \right \rVert / \eps}{
  \left \lVert \bm{\alpha} \right \rVert}\right).
\end{equation}
It's possible to rewrite \(\sum_j \delta_j \phi_j\left(\bm{\alpha}\right) =
F_{\bm{c}} \bm{\delta}\) where \(F_{\bm{c}} =
\left[\frac{\partial F_i}{\partial c_j}\right]_{ij} =
\left[\begin{array}{c c c} \phi_0\left(\bm{\alpha}\right) & \cdots &
\phi_n\left(\bm{\alpha}\right) \end{array}\right]\) is the Jacobian of \(F\)
with respect to the coefficients \(\bm{c}\). With this representation, the
Higham condition number has a closed form since
\begin{equation}
  \left \lVert
  J^{-1} F_{\bm{c}} \bm{\delta} \right \rVert / \eps \leq
  \left \lVert J^{-1} F_{\bm{c}} \right \rVert \left \lVert \bm{\delta} \right
  \rVert / \eps \leq \left \lVert J^{-1} F_{\bm{c}} \right \rVert \left \lVert
  \bm{c} \right \rVert
\end{equation}
for any matrix norm that is compatible with the vector norm used on
\(\bm{\delta}\). The Frobenius matrix norm and vector 2-norm can be
combined to give a condition number that is straightforward to compute:
\begin{equation}\label{eq:higham-cond-num}
  \kappa_H = \left \lVert J^{-1} F_{\bm{c}} \right \rVert_F
  \frac{\left \lVert \bm{c}
  \right \rVert_2}{\left \lVert \bm{\alpha} \right \rVert_2}.
\end{equation}

@@

\begin{proof}
Let the curve \(b_0(s)\) be of degree \(m\) and \(b_1(t)\) be of degree \(n\).
Then \(F(s, t)\) can be written as \\ \(2(m + 1) + 2(n + 1)\) terms:
\begin{equation}\label{eq:bezier-full-basis}
F(s, t) =
  \sum_{i = 0}^m c_i^{(1)} \left[ \begin{array}{c}
  B_{i, m}(s) \\ 0 \end{array}\right] +
  \sum_{i = 0}^m c_i^{(2)} \left[ \begin{array}{c}
  0 \\ B_{i, m}(s) \end{array}\right] +
  \sum_{j = 0}^n c_j^{(3)} \left[ \begin{array}{c}
  -B_{j, n}(t) \\ 0 \end{array}\right] +
  \sum_{j = 0}^n c_j^{(4)} \left[ \begin{array}{c}
  0 \\ -B_{j, n}(t) \end{array}\right].
\end{equation}
Since \(F(s, t) = b_0(s) - b_1(t)\) we have Jacobian \(J(s, t) =
\left[ \begin{array}{c c} b_0'(s) & -b_1'(t) \end{array}\right]\). Since we
are considering a transversal intersection, we have
\(\det J(\alpha, \beta) \neq 0\). In a perturbation, we replace each
\(c_j^{(k)}\) with a \(\delta_j^{(k)}\) bounded by
\(\left|\delta_j^{(k)}\right| \leq \eps \left|c_j^{(k)}\right|\).
By writing \(J^{-1} = \left[ \begin{array}{c c}
\bm{v} & \bm{w} \end{array}\right]\), we have
\begin{multline}
J^{-1}\left(\bm{\alpha}\right) \sum_j \delta_j
  \phi_j\left(\bm{\alpha}\right) =
  \left[\sum_{i = 0}^m \delta_i^{(1)} B_{i, m}(\alpha) +
  \sum_{j = 0}^n \delta_j^{(3)} B_{j, n}(\beta)\right] \bm{v} \\
  + \left[\sum_{i = 0}^m \delta_i^{(2)} B_{i, m}(\alpha) +
  \sum_{j = 0}^n \delta_j^{(4)} B_{j, n}(\beta)\right] \bm{w} =
  \nu_1 \bm{v} + \nu_2 \bm{w}
\end{multline}
where
\begin{align}
\left|\nu_1\right| / \eps &\leq \sum_{i = 0}^m
  \left|c_{i}^{(1)}\right| B_{i, m}\left(\alpha\right) + \sum_{j = 0}^n
  \left|c_{j}^{(3)}\right| B_{j, n}\left(\beta\right) =
  \widetilde{x}_0(\alpha) + \widetilde{x}_1(\beta) = \mu_1 \\
\left|\nu_2\right| / \eps &\leq \sum_{i = 0}^m
  \left|c_{i}^{(2)}\right| B_{i, m}\left(\alpha\right) + \sum_{j = 0}^n
  \left|c_{j}^{(4)}\right| B_{j, n}\left(\beta\right) =
  \widetilde{y}_0(\alpha) + \widetilde{y}_1(\beta) = \mu_2.
\end{align}
As in \eqref{eq:tri-ineq}, the bound can be attained by choosing the
sign and magnitude of each perturbation so that
\(\delta_j^{(k)} B_{j, d} = \eps \left|c_j^{(k)}\right| B_{j, d}\).
The factor of \(\eps\) can be cancelled to give the relative root
condition number
\begin{align}
\kappa_{\alpha, \beta} &= \frac{1}{\sqrt{\alpha^2 + \beta^2}}
  \sup_{\left|\nu_k\right| \leq \mu_k} \left \lVert \nu_1 \bm{v} +
  \nu_2 \bm{w} \right \rVert_2 \\
  &=
  \sqrt{\frac{\sup_{\left|\nu_k\right| \leq \mu_k}
  \nu_1^2 \left(\bm{v} \cdot \bm{v}\right) +
  2 \nu_1 \nu_2 \left(\bm{v} \cdot \bm{w}\right) +
  \nu_2^2 \left(\bm{w} \cdot \bm{w}\right)}{\alpha^2 + \beta^2}}
  \label{eq:intersect-cond-num}.
\end{align}
Now we seek to maximize the objective function \(\theta(\nu_1, \nu_2) =
\nu_1^2 \left(\bm{v} \cdot \bm{v}\right) +
2 \nu_1 \nu_2 \left(\bm{v} \cdot \bm{w}\right) +
\nu_2^2 \left(\bm{w} \cdot \bm{w}\right)\) in the rectangle
\(\left[-\mu_1, \mu_1\right] \times \left[-\mu_2, \mu_2\right]\).

@@

Along the boundary of the rectangle,
we fix one of \(\nu_1\) or \(\nu_2\) and the resulting univariate function is
an up-opening parabola. For example, fixing \(\nu_2 = c\) gives
\(\theta(\nu_1, c) =
\nu_1^2 \left(\bm{v} \cdot \bm{v}\right) +
\nu_1\left[2 c \left(\bm{v} \cdot \bm{w}\right)\right] +
c^2 \left(\bm{w} \cdot \bm{w}\right)\) which has positive lead coefficient
\(\|\bm{v}\|_2^2\). The lead coefficient cannot be \(0\) since if \(\bm{v}\)
were the zero vector we'd have \(\det J = 0\).
Since \(\theta\) is an up-opening parabola along the boundary, any critical
point must be a local minimum.

@@

\subsection{Family of Lines Approaching Coincident}

@@

At the solution, we have \(\mu_1 = \mu_2 = 2\) and
\begin{equation}
J^{-1} = \frac{1}{r}
\left[ \begin{array}{c c} r & 1 \\ 0 & 1 \end{array}\right]
\end{equation}
so that \(\bm{v} \cdot \bm{v} = 1\), \(\bm{v} \cdot \bm{w} =
1/r\) and \(\bm{w} \cdot \bm{w} = 2/r^2\). Again have a condition number
\begin{equation}
\kappa_{\alpha, \beta} = \sqrt{\frac{4}{r^2} + \frac{4}{r} + 2} =
  \frac{2}{r} + 1 + \frac{r}{4} + \bigO{r^2}.
\end{equation}
that increases towards infinity as the parameter \(r \longrightarrow 0^+\).

>>>>

Consider a smooth function \(F: \reals^n \longrightarrow \reals^n\)
with Jacobian \(F_{\bm{x}} = J\). We want to consider a special class of
functions of the form \(F\left(\bm{x}\right) = \sum_j c_j
\phi_j\left(\bm{x}\right)\) where the basis
functions \(\phi_j\) are also smooth functions
\(\reals^n \longrightarrow \reals^n\)
and each \(c_j \in \reals\). We want to consider the effects on a root
\(\bm{\alpha} \in \reals^n\) of a perturbation in one of the
coefficients \(c_j\). We examine the perturbed function
\begin{equation}
G\left(\bm{x}, \delta\right) = F\left(\bm{x}\right) +
\delta \phi_j\left(\bm{x}\right).
\end{equation}
Since \(G\left(\bm{\alpha}, 0\right) = \bm{0}\), if \(J^{-1}\) exists at
\(\bm{x} = \bm{\alpha}\),
the implicit function theorem tells us that we can define
\(\bm{x}\) via
\begin{equation}
G\left(\bm{x}\left(\delta\right), \delta\right) = \bm{0}.
\end{equation}
Taking the derivative with respect to \(\delta\) we find that
\(\bm{0} = G_{\bm{x}} \bm{x}' + G_{\delta}\). Plugging in
\(\delta = 0\) we find that \(\bm{0} = J\left(\bm{\alpha}\right) \bm{x}' +
\phi_j\left(\bm{\alpha}\right)\), hence we
conclude that
\begin{equation}
\bm{x}\left(\delta\right) = \bm{\alpha} - J^{-1}\left(\bm{\alpha}\right)
  \phi_j\left(\bm{\alpha}\right) \delta + \bigO{\delta^2}.
\end{equation}
This gives
\begin{equation}
\frac{\left \lVert J^{-1}\left(\bm{\alpha}\right)
  \phi_j\left(\bm{\alpha}\right) \right \rVert}{
  \left \lVert \bm{\alpha} \right \rVert}.
\end{equation}
as the relative condition number for a perturbation in \(c_j\).
By considering perturbations in \emph{all} of the coefficients:
\(\left|\delta_j\right| \leq \eps \left|c_j\right|\), a similar analysis
gives a root function
\begin{equation}
\bm{x}\left(\delta_0, \ldots, \delta_n\right) = \bm{\alpha} -
  J^{-1}\left(\bm{\alpha}\right) \sum_{j = 0}^n \delta_j
  \phi_j\left(\bm{\alpha}\right) + \bigO{\eps^2}.
\end{equation}
With this, we can define a root condition number

@@

In \cite[Chapter~25, Section 25.4]{Higham2002} a similar definition is
given. Instead of bounding the perturbations component-wise, it bounds
the entire perturbation vector \(\bm{\delta}\)
\begin{equation}
  \lim_{\eps \to 0} \left(\sup_{\left \lVert \bm{\delta} \right \rVert \leq
  \eps \left \lVert \bm{c} \right \rVert} \frac{\left \lVert
  J^{-1}\left(\bm{\alpha}\right) \sum_j \delta_j
  \phi_j\left(\bm{\alpha}\right) \right \rVert / \eps}{
  \left \lVert \bm{\alpha} \right \rVert}\right).
\end{equation}
It is possible to rewrite \(\sum_j \delta_j \phi_j\left(\bm{\alpha}\right) =
F_{\bm{c}} \bm{\delta}\) where \(F_{\bm{c}} =
\left[\frac{\partial F_i}{\partial c_j}\right]_{ij} =
\left[\begin{array}{c c c} \phi_0\left(\bm{\alpha}\right) & \cdots &
\phi_n\left(\bm{\alpha}\right) \end{array}\right]\) is the Jacobian of \(F\)
with respect to the coefficients \(\bm{c}\). With this representation, the
Higham condition number has a closed form since
\begin{equation}
  \left \lVert
  J^{-1} F_{\bm{c}} \bm{\delta} \right \rVert / \eps \leq
  \left \lVert J^{-1} F_{\bm{c}} \right \rVert \left \lVert \bm{\delta} \right
  \rVert / \eps \leq \left \lVert J^{-1} F_{\bm{c}} \right \rVert \left \lVert
  \bm{c} \right \rVert
\end{equation}
for any matrix norm that is compatible with the vector norm used on
\(\bm{\delta}\). The Frobenius matrix norm and vector 2-norm can be
combined to give a condition number that is straightforward to compute:
\begin{equation}\label{eq:higham-cond-num}
  \kappa_H = \left \lVert J^{-1} F_{\bm{c}} \right \rVert_F
  \frac{\left \lVert \bm{c}
  \right \rVert_2}{\left \lVert \bm{\alpha} \right \rVert_2}.
\end{equation}

@@

\begin{proof}
Let the curve \(b_0(s)\) be of degree \(m\) and \(b_1(t)\) be of degree \(n\).
Then \(F(s, t)\) can be written as a sum of \(2(m + 1) + 2(n + 1)\) terms:
\begin{equation}\label{eq:bezier-full-basis}
F(s, t) =
  \sum_{i = 0}^m c_i^{(1)} \left[ \begin{array}{c}
  B_{i, m}(s) \\ 0 \end{array}\right] +
  \sum_{i = 0}^m c_i^{(2)} \left[ \begin{array}{c}
  0 \\ B_{i, m}(s) \end{array}\right] +
  \sum_{j = 0}^n c_j^{(3)} \left[ \begin{array}{c}
  -B_{j, n}(t) \\ 0 \end{array}\right] +
  \sum_{j = 0}^n c_j^{(4)} \left[ \begin{array}{c}
  0 \\ -B_{j, n}(t) \end{array}\right].
\end{equation}
Since \(F(s, t) = b_0(s) - b_1(t)\) we have Jacobian \(J(s, t) =
\left[ \begin{array}{c c} b_0'(s) & -b_1'(t) \end{array}\right]\). Since we
are considering a transversal intersection, we have
\(\det J(\alpha, \beta) \neq 0\). In a perturbation, we replace each
\(c_j^{(k)}\) with a \(\delta_j^{(k)}\) bounded by
\(\left|\delta_j^{(k)}\right| \leq \eps \left|c_j^{(k)}\right|\).
By writing \(J^{-1} = \left[ \begin{array}{c c}
\bm{v} & \bm{w} \end{array}\right]\), we have
\begin{multline}
J^{-1}\left(\bm{\alpha}\right) \sum_j \delta_j
  \phi_j\left(\bm{\alpha}\right) =
  \left[\sum_{i = 0}^m \delta_i^{(1)} B_{i, m}(\alpha) +
  \sum_{j = 0}^n \delta_j^{(3)} B_{j, n}(\beta)\right] \bm{v} \\
  + \left[\sum_{i = 0}^m \delta_i^{(2)} B_{i, m}(\alpha) +
  \sum_{j = 0}^n \delta_j^{(4)} B_{j, n}(\beta)\right] \bm{w} =
  \nu_1 \bm{v} + \nu_2 \bm{w}
\end{multline}
where
\begin{align}
\left|\nu_1\right| / \eps &\leq \sum_{i = 0}^m
  \left|c_{i}^{(1)}\right| B_{i, m}\left(\alpha\right) + \sum_{j = 0}^n
  \left|c_{j}^{(3)}\right| B_{j, n}\left(\beta\right) =
  \widetilde{x}_0(\alpha) + \widetilde{x}_1(\beta) = \mu_1 \\
\left|\nu_2\right| / \eps &\leq \sum_{i = 0}^m
  \left|c_{i}^{(2)}\right| B_{i, m}\left(\alpha\right) + \sum_{j = 0}^n
  \left|c_{j}^{(4)}\right| B_{j, n}\left(\beta\right) =
  \widetilde{y}_0(\alpha) + \widetilde{y}_1(\beta) = \mu_2.
\end{align}
As in \eqref{eq:tri-ineq}, the bound can be attained by choosing the
sign and magnitude of each perturbation so that
\(\delta_j^{(k)} B_{j, d} = \eps \left|c_j^{(k)}\right| B_{j, d}\).
The factor \(\eps\) can be cancelled to give the relative root
condition number
\begin{align}
\kappa_{\alpha, \beta} &= \frac{1}{\sqrt{\alpha^2 + \beta^2}}
  \sup_{\left|\nu_k\right| \leq \mu_k} \left \lVert \nu_1 \bm{v} +
  \nu_2 \bm{w} \right \rVert_2 \\
  &=
  \sqrt{\frac{\sup_{\left|\nu_k\right| \leq \mu_k}
  \nu_1^2 \left(\bm{v} \cdot \bm{v}\right) +
  2 \nu_1 \nu_2 \left(\bm{v} \cdot \bm{w}\right) +
  \nu_2^2 \left(\bm{w} \cdot \bm{w}\right)}{\alpha^2 + \beta^2}}
  \label{eq:intersect-cond-num}.
\end{align}
Now we seek to maximize the objective function \(\theta(\nu_1, \nu_2) =
\nu_1^2 \left(\bm{v} \cdot \bm{v}\right) +
2 \nu_1 \nu_2 \left(\bm{v} \cdot \bm{w}\right) +
\nu_2^2 \left(\bm{w} \cdot \bm{w}\right)\) in the rectangle
\(\left[-\mu_1, \mu_1\right] \times \left[-\mu_2, \mu_2\right]\).

@@

Along the boundary of the rectangle,
we fix one of \(\nu_1\) or \(\nu_2\) and the resulting univariate function is
an up-opening parabola. For example, fixing \(\nu_2 = c\) gives
\(\theta(\nu_1, c) =
\nu_1^2 \left(\bm{v} \cdot \bm{v}\right) +
\nu_1\left[2 c \left(\bm{v} \cdot \bm{w}\right)\right] +
c^2 \left(\bm{w} \cdot \bm{w}\right)\) which has positive lead coefficient
\(\|\bm{v}\|_2^2\). The lead coefficient cannot be \(0\) since if \(\bm{v}\)
were the zero vector we would have \(\det J = 0\).
Since \(\theta\) is an up-opening parabola along the boundary, any critical
point must be a local minimum.

@@

\subsection{Family of Lines Approaching Coincidence}

@@

At the solution, we have \(\mu_1 = \mu_2 = 2\) and
\begin{equation}
J^{-1} = \frac{1}{r}
\left[ \begin{array}{c c} r & 1 \\ 0 & 1 \end{array}\right]
\end{equation}
so that \(\bm{v} \cdot \bm{v} = 1\), \(\bm{v} \cdot \bm{w} =
1/r\) and \(\bm{w} \cdot \bm{w} = 2/r^2\). Again we have a condition number
\begin{equation}
\kappa_{\alpha, \beta} = \sqrt{\frac{4}{r^2} + \frac{4}{r} + 2} =
  \frac{2}{r} + 1 + \frac{r}{4} + \bigO{r^2}.
\end{equation}
that increases towards infinity as the parameter \(r \longrightarrow 0^+\).
